Cluster-efficient pangenome graph construction with nf-core/pangenome

### 2024-10-08 manuscript/main.tex
reviewer2_2_long

	The pipeline begins with an all-to-all alignment of the input sequences using the whole-chromosome pairwise sequence aligner WFMASH \citep{wfmash}, avoiding reference, order, or orientation bias, allowing every sequence to serve as a reference.
	%This avoids reference, order, or orientation bias, and allows each sequence in the pangenome to serve as a reference when exploring related variation. 
	In the pangenome graph induction step SEQWISH \citep{Garrison2022}, an alignment to variation graph inducer, converts the sequence alignments into a variation graph. 
	%SEQWISH can recapture homology relationships that may not be encoded in the initial alignments. 
	%This allows us to randomly sparsify the mappings computed by WFMASH. 
	%The applied heuristic is based on the Erdös-Rényi random graph model \citep{Bollobs2001}. 
	The graph is then simplified using SMOOTHXG \citep{Garrison2023}:
	A 1-dimensional (1D) graph embedding \citep{Heumos2023} orders the graphs’ nodes to best-match the nucleotide distances of the genomic paths of the graph.
	%This considers the local spurious complexity inherent in the raw pairwise alignments. 
	%If ignored, this could lead to difficulties in downstream processing of the graph.
	%Therefore, SMOOTHXG iteratively applies a local Multiple Sequence Alignment (MSA), partial order alignment (POA) \citep{Lee2002}, to refine and compress the pangenome graph. A 1-dimensional (1D) graph embedding \citep{Heumos2023} gives rise to the regions the kernel is applied to. 
	%After applying the graph embedding algorithm, the graphs’ node order best-matches the nucleotide distances of the genomic paths of the graph. 
	Next, the graph is split into partially overlapping segments. 
	%The sequences of each segment are realigned with POA. 
	The sequences of each segment are realigned with a local Multiple Sequence Alignment (MSA) kernel, partial order alignment (POA) \citep{Lee2002}.
	Afterwards, the segments are laced back together into a variation graph. 
	By default, the SMOOTHXG process is applied 3 times in order to smoothen the edge effects at the boundaries of the segments.
	Finally, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph.
	%In the final normalization phase, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph. 
	%Subsequently, the 1D graph embedding algorithm is reapplied to ensure a human-readable representation of the processed data. 
	%For more details about each step, we refer to the original PGGB paper (Garrison et al. 2023).

@@

	The Human Pangenome Resource Consortium (HPRC) recently built a draft human pangenome reference of 90 haplotypes \citep{Liao2023}. 
	However, haplotype data for thousands of individuals already exists generated by the 1000 Genomes Project (1KGP) \citep{Durbin2010}. 
	As a use case study, we used nf-core/pangenome to build a pangenome graph of 1000 chromosome 19 haplotypes \citep{Kuhnle2020} within 3 days. 
	The CO2e was 22.52 kg. PGGB built the same graph within 7 days. 
	In Fig.~\ref{fig:hist_chr19} the pangenome growth curve generated with PANACUS \citep{Liao2023} shows a growth of the number of nucleotides with an increasing number of haplotypes.
	The size of the softcore pangenome does not change with increasing numbers of haplotypes.
	%One reason for this could be that the input sequences were generated from short read data leaving out complex regions like the centromere.

@@

	To evaluate the pipeline’s scalability, we built a pangenome graph of 2146 \textit{E. coli} sequences. 
	nf-core/pangenome built the pangenome graph in 10 days, emitting 175.18 kg of CO2e. 
	Due to wall clock time restrictions on our cluster, PGGB was not able to finish the graph construction within 30 days. 
	To build a reasonable pangenome growth curve (Fig.~\ref{fig:hist_ecoli}) we dropped all paths containing “plasmid” (130 in total) in their name. 
	The softcore pangenome of the graph does not change with an increasing number of haplotypes (stable at ~3Mb of sequence), but the general growth curve is steep. 
	%One reason for such a high pangenomic growth is most likely horizontal gene transfer: 
	%Bacteria incorporate genes from each other at different locations in their genome. Other reasons could be sequencing errors or human contamination \citep{Breitwieser2019}.

@@

	Although, we expect our pipeline to scale well for future pangenome graph construction challenges, such as for the next HPRC phase which targets 350 individuals, there still is potential for further optimization: 
	IMplicit Pangenome Graph (IMPG) (\href{https://github.com/ekg/impg}{https://github.com/ekg/impg}) which is a tool that extracts homologous loci from all genomes mapped to a specific target region. 
	This would allow us to break the whole genome multiple alignments into smaller pieces, construct a pangenome graph for each piece, and lace these together into a full graph with \href{https://github.com/pangenome/gfalace}{https://github.com/pangenome/gfalace}. 
	%Our first trials 
	%(\href{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}) 
	%suggest following this direction.

	We anticipate the pipeline, or its parts, will enhance current single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. Looking ahead, pangenome construction pipelines like nf-core/pangenome will play a pivotal role in studying entire populations, single-cell whole genome sequencing analysis, and constructing personalized (medical) pangenome references \citep{Sirn2023}.
	%We believe that the pipeline or parts of it will enhance existing single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. We envision for the future that pangenome construction pipelines, like nf-core/pangenome, will play a pivotal role in studying whole populations, single-cell whole genome sequencing analysis, and, for example, specifically when it comes to the construction of personalized (medical) pangenome references \citep{Sirn2023}. 
	\vspace{-0.6cm}
	\section*{Software and data availability}

@@

	The FASTA of the chromosome 19 data set was downloaded in December 2023 from \href{http://dolomit.cs.tu-dortmund.de/chr19.1000.fa.xz}{http://dolomit.cs.tu-dortmund.de/chr19.1000.fa.xz}. The data set is described in \citep{Kuhnle2020}. Statistics of the built pangenome graph can be seen in Supplementary Fig. ~\ref{fig:chr19_multiqc}. The initial graph contains over 97\% of Ns. We applied \textit{odgi crush} (odgi version 0.8.6), which crushes consecutive Ns of all nodes containing Ns into just one N per node, to the 1KGP chromosome 19 pangenome graph. This brings down the number of Ns from over 3B to exactly 6000 (Suppl. Fig.~\ref{fig:chr19_multiqc}). 
	The 2D visualization (Suppl. Fig.~\ref{fig:chr19_2D}) is perfectly linear without any large SVs present hinting that only short-read data was used to create the haplotype sequences.
	In contrast, the 2D layout of the HPRC PGGB chromosome 19 pangenome graph \citep{Heumos2023} clearly presents SVs, especially in the centromere's location. Investigating these complex regions of a human chromosome is only possible when using long-read assemblies for graph construction.

>>>>

	The pipeline begins with an all-to-all alignment of the input sequences using the whole-chromosome pairwise sequence aligner WFMASH \citep{wfmash}, avoiding reference, order, or orientation bias, allowing every sequence to serve as a reference.
	%This avoids reference, order, or orientation bias, and allows each sequence in the pangenome to serve as a reference when exploring related variation. 
	In the pangenome graph induction step SEQWISH \citep{Garrison2022}, an alignment to variation graph inducer, converts the sequence alignments into a variation graph. 
	%SEQWISH can recapture homology relationships that may not be encoded in the initial alignments. 
	%This allows us to randomly sparsify the mappings computed by WFMASH. 
	%The applied heuristic is based on the Erdös-Rényi random graph model \citep{Bollobs2001}. 
	The graph is then simplified using SMOOTHXG \citep{Garrison2023}:
	A 1-dimensional (1D) graph embedding \citep{Heumos2024} orders the \red{graph’s} nodes to \red{best match} the nucleotide distances of the genomic paths of the graph.
	%This considers the local spurious complexity inherent in the raw pairwise alignments. 
	%If ignored, this could lead to difficulties in downstream processing of the graph.
	%Therefore, SMOOTHXG iteratively applies a local Multiple Sequence Alignment (MSA), partial order alignment (POA) \citep{Lee2002}, to refine and compress the pangenome graph. A 1-dimensional (1D) graph embedding \citep{Heumos2023} gives rise to the regions the kernel is applied to. 
	%After applying the graph embedding algorithm, the graphs’ node order best-matches the nucleotide distances of the genomic paths of the graph. 
	Next, the graph is split into partially overlapping segments. 
	%The sequences of each segment are realigned with POA. 
	The sequences of each segment are realigned with a local Multiple Sequence Alignment (MSA) kernel, partial order alignment (POA) \citep{Lee2002}.
	Afterwards, the segments are laced back together into a variation graph. 
	By default, the SMOOTHXG process is applied 3 times in order to \red{smooth} the edge effects at the boundaries of the segments.
	Finally, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph.
	%In the final normalization phase, we employ GFAFFIX \citep{Liao2023} to systematically condense redundant nodes within the graph. 
	%Subsequently, the 1D graph embedding algorithm is reapplied to ensure a human-readable representation of the processed data. 
	%For more details about each step, we refer to the original PGGB paper (Garrison et al. 2023).

@@

	\red{The Human Pangenome Resource Consortium (HPRC) recently built a draft human pangenome of 90 haplotypes \citep{Liao2023}. 
	However, haplotype data for thousands of individuals already exists, generated by the 1000 Genomes Project (1KGP) \citep{Durbin2010}. 
	As a use case, we used nf-core/pangenome to build a graph of 1000 chromosome 19 haplotypes \citep{Kuhnle2020} in 3 days, emitting 22.52 kg CO2e.
	PGGB took 7 days for the same task. 
	In Fig.~\ref{fig:hist_chr19} the pangenome growth curve generated with PANACUS \citep{Liao2023} shows nucleotide growth as more haplotypes are added.}
	\red{The softcore pangenome, defined as sequences traversed by 95\% of haplotypes, represents the majority of the pangenome even with 1000 haplotypes. 
	This stability may be due to the exclusion of complex regions like the centromere in the short-read data.
	}

@@

	To evaluate the pipeline’s scalability, we built a pangenome graph of 2146 \textit{E. coli} sequences. 
\red{	The nf-core/pangenome graph was completed in 10 days, emitting 175.18 kg CO2e, while PGGB could not finish within 30 days due to cluster time restrictions.}
\red{	For the growth curve (Fig.~\ref{fig:hist_ecoli}) we excluded 130 plasmid sequences. 
	The softcore pangenome remains stable at ~3Mb, but its size constitutes less than 10\% of the built pangenome.}
	This substantial pangenomic growth is likely driven by horizontal gene transfer, as bacteria incorporate genes from one another at various genomic locations.
	Other reasons could be sequencing errors or human contamination \citep{Breitwieser2019}.

@@

	\red{Although} we expect our pipeline to scale well for future pangenome graph construction challenges, such as for the next HPRC phase which targets 350 individuals, there still is potential for further optimization: 
	IMplicit Pangenome Graph (IMPG) (\href{https://github.com/ekg/impg}{https://github.com/ekg/impg}) which is a tool that extracts homologous loci from all genomes mapped to a specific target region. 
	This would allow us to break the whole genome multiple alignments into smaller pieces, construct a pangenome graph for each piece, and lace these together into a full graph with \href{https://github.com/pangenome/gfalace}{https://github.com/pangenome/gfalace}. 
	%Our first trials 
	%(\href{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}{https://hackmd.io/vh0Ap65uQTm40DdleA-Oow}) 
	%suggest following this direction.

	We anticipate the pipeline, or its parts, will enhance current single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. Looking ahead, pangenome construction pipelines like nf-core/pangenome will play a pivotal role in studying entire populations, single-cell whole genome sequencing analysis, and constructing personalized (medical) pangenome references \citep{Sirn2024}.
	%We believe that the pipeline or parts of it will enhance existing single linear reference analysis methods to explore whole population variation instead of focusing on one reference only. We envision for the future that pangenome construction pipelines, like nf-core/pangenome, will play a pivotal role in studying whole populations, single-cell whole genome sequencing analysis, and, for example, specifically when it comes to the construction of personalized (medical) pangenome references \citep{Sirn2024}. 
	\vspace{-0.6cm}
	\section*{Software and data availability}

@@

	The FASTA of the chromosome 19 data set was downloaded in December 2023 from \href{http://dolomit.cs.tu-dortmund.de/chr19.1000.fa.xz}{http://dolomit.cs.tu-dortmund.de/chr19.1000.fa.xz}. The data set is described in \citep{Kuhnle2020}. Statistics of the built pangenome graph can be seen in Supplementary Fig. ~\ref{fig:chr19_multiqc}. The initial graph contains over 97\% of Ns. We applied \textit{odgi crush} (odgi version 0.8.6), which crushes consecutive Ns of all nodes containing Ns into just one N per node, to the 1KGP chromosome 19 pangenome graph. This brings down the number of Ns from over 3B to exactly 6000 (Suppl. Fig.~\ref{fig:chr19_multiqc}). 
	The 2D visualization (Suppl. Fig.~\ref{fig:chr19_2D}) is perfectly linear without any large SVs present hinting that only short-read data was used to create the haplotype sequences.
	In contrast, the 2D layout of the HPRC PGGB chromosome 19 pangenome graph \citep{Heumos2024} clearly presents SVs, especially in the centromere's location. Investigating these complex regions of a human chromosome is only possible when using long-read assemblies for graph construction.
